---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mmatrix

<!-- badges: start -->

<!-- badges: end -->

mmatrix provides a safe, sparse, and reproducible formula-to-matrix pipeline for high-dimensional data. It extends R’s base model.matrix() workflow by offering transparent collinearity handling, stable factor encoding, and detailed reporting of dropped or aliased columns. The package centers on a main constructor that builds a sparse design matrix and a frozen encoding specification (“spec”) containing factor levels, contrasts, and transformation rules for consistent application to new data.

## Installation

You can install the development version of mmatrix from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gykwon-glitch/mmatrix")
```

## TO DO

1. Finish making the code for mm_predict, considering using S3 class print instead of mm_report.
2. (if time permits) do several tests or make vignettes

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(mmatrix)
## basic example code
```
